Forecasting the 2024 U.S. Presidential Election: A Poll-of-Polls Approach for Predicting the Outcome

### 2024-11-04 paper/paper.qmd
first model draft

```{r}
# Define the binary response variable if not already defined
# Create a binary `win` variable where pct > 50% indicates a "win"
data <- data %>%
  mutate(win = ifelse(pct > 50, 1, 0))
# Fit the full logistic regression model with all predictors
full_model <- glm(
  win ~ sample_size + pollscore + transparency_score + 
          methodology + start_date + end_date + state + party, 
  data = data, 
  family = binomial
)
# Use stepwise selection based on BIC
best_model <- stepAIC(
  full_model, 
  direction = "both", 
  k = log(nrow(data)) # Set k to log(n) for BIC
)

# Display the summary of the selected model based on BIC
summary(best_model)
# View final model formula and BIC value
formula(best_model)
BIC(best_model)
```

#model

# Model Analysis

In this study, we employed a Generalized Linear Model (GLM) to forecast the winner of the upcoming US presidential election. The model uses polling data to estimate the probability of each candidate winning a given state, enabling us to predict the overall election result. We used a logistic regression approach, where the outcome is a binary variable (1 if the candidate wins, 0 otherwise).

## Model Set-Up

The model takes the following form:

$$
P(y_i = 1) = \text{logit}^{-1}(\alpha + \beta_1 \cdot \text{pollscore}_i + \beta_2 \cdot \text{log\_sample\_size}_i + \gamma_{\text{state}_i} + \delta_{\text{pollster\_rating}_i})
$$

where (y_i) represents the probability of candidate (i) winning, ( \alpha ) is the intercept, and the coefficients (( \beta, \gamma, \delta )) represent the effects of predictors like poll score, sample size, state, and pollster rating.

The model was trained using polling data, aggregated from multiple sources (the poll-of-polls approach), which helped to minimize biases associated with individual polls. The logistic link function was employed to model the binary outcome of winning or losing in each state.

## Election Probability Estimation and Winner Prediction

The GLM provides a probability estimate for each candidate in every state. We applied a threshold of 0.5 to classify a candidate as the winner for each state. By aggregating the predicted winners across all states, we determined the overall forecast for the election.

The model results suggest that Candidate A has a predicted win in several key battleground states, giving them an estimated advantage in the electoral college. Based on these state-level predictions, Candidate A is forecasted to win the upcoming US presidential election.

## Model Diagnostics

We performed several diagnostic evaluations to ensure the reliability of our model:

-   **Residual Analysis**: Examined residual vs. fitted plots to assess any systematic patterns.
-   **Normal Q-Q Plot**: Used to verify that the residuals follow a normal distribution.
-   **Cook's Distance**: Identified potential influential observations that could have a disproportionate effect on the model.

These diagnostics indicated that the model performed well, with no significant violations of assumptions that would undermine the predictions.

## Model Results and Visualizations

The model coefficients and summary statistics are presented in Table @tbl-modelresults. These results provide insights into how different factors—such as poll quality and sample size—affect the probability of winning.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

library(rstanarm)

# Load the election model
election_model <- readRDS(file = here::here("models/election_model.rds"))
```
```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Model results for predicting the winner of the election"
#| warning: false

modelsummary::modelsummary(
  list(
    "Election Model" = election_model
  ),
  statistic = "mad",
  fmt = 2
)
```

@@

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

@@

Please don't use these as sub-heading labels - change them to be what your point actually is.

@@

Weaknesses and next steps should also be included.

@@

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

pp_check(first_model) +
  theme_classic() +
  theme(legend.position = "bottom")

posterior_vs_prior(first_model) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()
```

@@

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

plot(first_model, "trace")

plot(first_model, "rhat")
```

>>>>

# Model

# Description of the Model

The model used for this analysis is a generalized linear model (GLM) with a logistic regression link function, designed to predict the probability of a candidate winning an election based on polling data. The model takes into consideration several key predictors that may influence the likelihood of a candidate winning, including poll score, sample size, candidate party, and poll percentage. Specifically, the formula used in the model is:

$$
\text{logit}(P(Y = 1)) = \beta_0 + \beta_1 \cdot \text{pollscore} + \beta_2 \cdot \text{sample\_size} + \beta_3 \cdot \text{party\_binary} + \beta_4 \cdot \text{pct}
$$

where:
 - Y = 1 represents the outcome that a candidate wins the poll.
 - pollscore refers to the quality score assigned to the poll, reflecting its reliability.
 - sample_size represents the number of respondents in the poll.
 - party_binary is a binary indicator of the candidate's party (1 for Democratic, 0 for others).
 - pct represents the percentage of support the candidate received in the poll.

```{r}
# Load necessary libraries
library(rstanarm)
library(tidyverse)

# Load the cleaned data
polls_data_cleaned <- read_csv("../data/02-analysis_data/cleaned_president_polls.csv")

# Fit the logistic regression model
win_model <- stan_glm(
  formula = win ~ pollscore + sample_size + party_binary + pct,
  data = polls_data_cleaned,
  family = binomial(link = "logit"),
  prior = normal(location = 0, scale = 2.5, autoscale = TRUE),
  prior_intercept = normal(location = 0, scale = 2.5, autoscale = TRUE),
  seed = 853
)

# Print summary of the model
summary(win_model)
```
## Model Performance and Interpretation
The model's output includes estimates for each predictor, providing insights into their relative importance in predicting a candidate's probability of winning. The summary of the model reveals the following key points:

Poll Score (pollscore): The coefficient for pollscore indicates the extent to which the quality of a poll influences the predicted probability of a candidate winning. Higher-quality polls are expected to provide more reliable estimates, and therefore have a stronger effect on the predicted outcome.

Sample Size (sample_size): The coefficient for sample_size shows the influence of the number of respondents on the model. Polls with larger sample sizes are generally more reliable, contributing positively to the accuracy of predictions.

Party (party_binary): The party_binary variable highlights any potential advantage or disadvantage linked to party affiliation, with 1 representing Democratic candidates. This variable helps capture any systemic biases or influences based on party identification.

Poll Percentage (pct): The pct variable is a key predictor as it directly measures the level of support a candidate receives. A higher percentage is expected to lead to a higher probability of winning.

```{r}
# Calculate predicted win probabilities
polls_data_cleaned <- polls_data_cleaned %>%
  mutate(predicted_win_prob = predict(win_model, type = "response"))

# Display a few rows of data with predicted probabilities
polls_data_cleaned %>%
  dplyr::select(state, candidate_name, pct, predicted_win_prob) %>%
  head()

# Save the model and predictions for later use
saveRDS(win_model, "../models/election_win_model.rds")
write_csv(polls_data_cleaned, "../data/04-model_results/polls_with_predictions.csv")
```
The model described above allows us to predict the likelihood of each candidate winning based on the given polling data, using a logistic regression approach. The model was implemented using rstanarm to provide a Bayesian estimation, which helps incorporate uncertainty into our predictions.

@@

@@

@@

@@

```{r}

```

@@

```{r}

```
